---
title: "ROSMAP_ProteomicsTMT_04.signature 2.comparison"
author: ""
date:   "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: united
    code_folding: show
    toc: true
#    style: BS831.css  ## feel free to remove this if you don't want it
    toc_float: true
---

```{r setup, include=FALSE}
## include these options to avoid unnecessarily verbose output
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

# Settings
```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(pheatmap))
suppressPackageStartupMessages(library(GSVA))
suppressPackageStartupMessages(library(org.Hs.eg.db))
suppressPackageStartupMessages(library(limma))
suppressPackageStartupMessages(library(edgeR))
suppressPackageStartupMessages(library(ComplexHeatmap))
suppressPackageStartupMessages(library(RColorBrewer))
suppressPackageStartupMessages(library(impute))
suppressPackageStartupMessages(library(hypeR))
suppressPackageStartupMessages(library(psych))
```

# Step 1: read in the cleaned Expression Set object from 01
```{r}
es1<- readRDS('./data/proteomics_QCed_es1.rds')

# some gene symbols have multiple entries
c(total=nrow(es1),unique=length(unique(fData(es1)$gene)))

# we 'uniquefy' by taking the replicate w/ larges1t variation
SD <- apply(exprs(es1),1,sd,na.rm=TRUE)              # compute SD of each gene
es1 <- es1[order(SD,decreasing=TRUE),]             # sort genes1 by SD (larges1t to smalles1t)
es1 <- es1[match(unique(fData(es1)$gene), # pick first occurrence of each
                     fData(es1)$gene),]


featureNames(es1) <- fData(es1)$gene
c(total=nrow(es1),unique=length(unique(fData(es1)$gene)))

## probes w/ NA's 
sum(is.na(exprs(es1)))
```

# step 2: APOE Signature2 Projection on the ROSMAP Proteomics TMT Dataset

```{r}
apoeSIG <- list(APOEup = c("APOE", "CAMP", "APOL1", "APOC1", "LYZ",  "CETP", "MPO", "APOC4", "CTSG", "FSTL1", "DEFA3", "GC", "PLG", "KNG1"),
                APOEdn = c( "APOB", "CTBS", "SERPINA6", "LUM"))
```

```{r}
## let's check overlap of the signature with genes in the LOAD dataset
data.frame(
  original=sapply(apoeSIG,length),
  matching=sapply(lapply(apoeSIG,intersect,toupper(featureNames(es1))),length)
) %>% dplyr::mutate(delta=original-matching)
```

## Exercise 1: GSVA Projection

```{r}
gsvaPar1<- gsvaParam(exprData = exprs(es1), apoeSIG,use="na.rm")
gsvaPar1
```

```{r}
gsva.es1 <- gsva(gsvaPar1, verbose=FALSE)
dim(gsva.es1)
```


```{r}
## adding combined up-dn scores
gsva.es1 <- rbind(gsva.es1,
                  APOEscore=gsva.es1["APOEup",]-gsva.es1["APOEdn",])
## show few entries    
print(gsva.es1[,1:5])
```
## Exercise 2: APOE Signature Projection Heatmap

```{r}
## just checking
all(sampleNames(es1)==colnames(gsva.es1))
```
```{r}
## plotting heatmap

annot <- pData(es1)[,"cogdx_1",drop=FALSE]
annotCol <- list(cogdx_1=c(AD="black",healthy="white"))

es.ord1 <- gsva.es1[,order(gsva.es1[nrow(gsva.es1),])]

color_pal <- colorRampPalette(c("blue", "white", "red"))(9)

pheatmap(es.ord1,
         color = color_pal,
         annotation_col = annot,
         annotation_colors = annotCol,
         cluster_rows=FALSE,
         cluster_cols=FALSE,
         show_rownames = TRUE,
         show_colnames = FALSE,
         scale = "row",height=0.25,
         main="APOE signature 2 ROSMAP Proteomics projection",
         cellheight=25)
```

## Test Association between APOE Signature Enrichment and LOAD Status

### method 1: Test Association by KS

```{r}
all(colnames(gsva.es1)==sampleNames(es1))
```
```{r}
apoe <- gsva.es1["APOEscore",]

apoe_scores_ad<- rank(apoe)[es1$cogdx_1=="AD"]
apoe_scores_ctrl<- rank(apoe)[es1$cogdx_1=="healthy"]
```

```{r}
# Perform the Kolmogorov-Smirnov test
ks_test_result <- ks.test(apoe_scores_ad, apoe_scores_ctrl)
print(ks_test_result)
```
```{r}
# plot 
# Calculate the CDF for both datasets
ad_cdf <- ecdf(apoe_scores_ad)
ctrl_cdf <- ecdf(apoe_scores_ctrl)

# Create a sequence of points that covers the range of both distributions
x_vals <- seq(min(c(apoe_scores_ad, apoe_scores_ctrl)), max(c(apoe_scores_ad, apoe_scores_ctrl)), length.out = 100)

# Calculate the CDF values for both distributions at each point
ad_cdf_vals <- ad_cdf(x_vals)
ctrl_cdf_vals <- ctrl_cdf(x_vals)

# Calculate the D statistic (the absolute difference between the CDFs)
d_values <- abs(ad_cdf_vals - ctrl_cdf_vals)

# Find the index and value of the maximum D statistic
max_d_index <- which.max(d_values)
max_d_value <- d_values[max_d_index]
max_d_x <- x_vals[max_d_index]

# Create a data frame for plotting
df <- data.frame(x = x_vals, D = d_values)

# Plot using ggplot2
library(ggplot2)
ggplot(df, aes(x = x, y = D)) +
  geom_line(color = "blue") +
  # Add a dashed vertical line at the midpoint of the x-axis
  geom_vline(xintercept = median(x_vals), linetype = "dashed", color = "black") +
  # Add a red dot at the max D value
  geom_point(aes(x = max_d_x, y = max_d_value), color = "red", size = 3) +
  # Label the max D value
  geom_text(aes(x = max_d_x, y = max_d_value, label = paste("Max D =", round(max_d_value, 3))),
            color = "red", vjust = -1, hjust = 1) +
  labs(title = "Kolmogorov-Smirnov D Statistic", x = "APOE Score", y = "D Statistic") +
  theme_minimal() +
  scale_x_continuous(limits = c(min(x_vals), max(x_vals))) +
  scale_y_continuous(limits = c(0, max(d_values)))

```

### method 2: Test Association by Differential Analysis
```{r}
# use lm
lmm <- lm(gsva.es1["APOEscore",] ~ es1$cogdx_1+es1$msex+es1$age_death_1+es1$pmi)
pval_cogdx <- summary(lmm)$coefficients["es1$cogdx_1AD", "Pr(>|t|)"]
beta<- summary(lmm)$coefficients["es1$cogdx_1AD", "Estimate"]

boxplot(gsva.es1["APOEscore",] ~ es1$cogdx_1,
        las=1,main="ROSMAP Proteomics", col=c('pink','lightblue'),pch="-",
        xlab=paste("lm pvalue =",round(pval_cogdx, 4),
                   ',',
                   'beta =', round(beta, 4)),
        ylab="APOE score")
```

```{r}
# use wilcox test
pval <- wilcox.test(gsva.es1["APOEscore",] ~ es1$cogdx_1)
beta<- mean(gsva.es1["APOEscore",colnames(es1)[es1$cogdx_1=='AD']]) - mean(gsva.es1["APOEscore",colnames(es1)[es1$cogdx_1=='healthy']])

boxplot(gsva.es1["APOEscore",] ~ es1$cogdx_1,
        las=1,main="ROSMAP Proteomics", col=c('pink','lightblue'),pch="-",
        xlab=paste("wilcox pvalue",round(pval$p.value, 4),
                   ',',
                   'beta =', round(beta, 4),
                   '\n(beta = mean (AD APOE score) - mean (healthy APOE score))'),
        ylab="APOE score")

```


# Step 3: Boxplots

```{r}
# read in limmaRes2 result
limmaRes2<- read.csv('./result/ROSMAP_TMT_DEP_limmaRes2.csv')
limmaRes2<- limmaRes2 %>% column_to_rownames(var = 'X')
```


```{r}
# boxplot of the signature proteins

ig <- unlist(apoeSIG, use.names = FALSE)

for (gene in ig) {
    if (gene %in% es1@featureData$gene) {
        # Extract expression values
        expr_values <- exprs(es1[es1@featureData$gene == gene, ])[1, ]
        
        # Group labels
        group_labels <- es1$cogdx_1
        
        # Perform Wilcoxon rank-sum test
        wilcox_pval <- wilcox.test(expr_values ~ group_labels)$p.value
        
        # Calculate Beta (Mean_AD - Mean_Healthy)
        mean_AD <- mean(expr_values[group_labels == "AD"], na.rm = TRUE)
        mean_Healthy <- mean(expr_values[group_labels == "healthy"], na.rm = TRUE)
        beta <- mean_AD - mean_Healthy  # Beta coefficient
        
        # Create boxplot
        boxplot(expr_values ~ group_labels,
                las = 1, main = paste0(gene, " ROSMAP Proteomics"), 
                col = c('pink', 'lightblue'), pch = "-",
                xlab = paste0(gene, 
                              ', ',
                              "limma2 padj = ", 
                              round(limmaRes2[limmaRes2$gene == gene, "adj.P.Val"], 4), 
                              ', ',
                              'log2FC = ', 
                              round(limmaRes2[limmaRes2$gene == gene, "logFC"], 4),
                              "\n",  # <-- Line break added here
                              'Wilcox p = ', 
                              signif(wilcox_pval, 4),
                              ', ',
                              'Beta = ', 
                              round(beta, 4)  # Show Beta coefficient
                ),
                ylab = paste0(gene, " z-scored log2 expression"))
    }
}
```

```{r}
# found two appearances of KNG1

limmaRes2 %>% filter(gene=='KNG1')
```

